The evolution of badges of status with learners

### 2019-07-01 QD_report.Rmd
Changes in the text of the Rmd, and adding the bibtex file with references.

---
title: "The evolution of badges of status with learners"
author: "Andrés Quiñones"
output: 
  pdf_document: 
    fig_caption: yes
    keep_tex: yes
header-includes: \usepackage{float}
                \floatplacement{figure}{H}
---

@@

```{r fig1, echo=FALSE, fig.cap="\\label{fig:HD_game}Hawk-dove game. Dashed line is the  game theoretical prediction for frequency of hawks.", out.width = '80%'}
knitr::include_graphics("Simulations//mutType_//basicHawkDove.png")
```

@@

To complete the receiver side of the signalling system, we introduce a new strategy in the game that we call _learner_. _Learners_ use a mixed strategy in the hawk-dove game. Which means, they play hawk (dove) with a certain probability.
Furthermore, the probability with which the play each strategy changes during their lifetime according to the payoff they recieve in previous interactions. In other words, _learners_ learn to behave in the game. To implement the learning
process, we use the reinforcement learning formalism. Specifically, we use the actor-critic algorithm. Individuals ( _learners_ ) perceive the _badge_ of their counterpart as the environmental state in which they find themselves. They estimate the value of that state and a preference for playing hawk (dove) in that state. They update their estimation and preference after each interaction they have. Given that the badge size is a real number between 0 and 1, there are infinitely many environmental states. Thus, for their value estimation and preference computation, individual must use a form of generalization. We let them
use as generalization a function approximation method based on radial basis functions. Each individual uses five _centers_ spread uniformly throughout the range of badge size variation (from 0 to 1). Each time an individual interacts with a conspecific the size of the badge of the interacting partner triggers are response from each one of the _centers_. This response decreases in strength as the badge size is further from the _center_. The decrease follows a radial basis function with gaussian shape. The sum of the responses given by all the centers provide the value estimation and the preference for one of the two pure strategies (hawk or dove). During the learning process individuals update the response that each one of these centers triggers. In the case of the preference for the strategies the numerical value resulting from the sum of the responses is translated to a probability of one of the two pure strategies through a logistic function. Figure \ref{fig:learning_cartoonRBF} shows a made up example of how the function approximation works for the actor (top panel - probability of choosing dove) and the critic (bottom panel - expected payoff from facing an individual with a given badge size).

@@

@@

@@

## How does learning look like in the Hawk-Dove game

@@

@@

@@

>>>>

---
title: "The evolution of badges of status with learners"
author: "Andrés Quiñones"
bibliography: SocialMorphCue.bib
output: 
  pdf_document: 
    fig_caption: yes
    keep_tex: yes
header-includes: \usepackage{float}
                \floatplacement{figure}{H}
---

@@

```{r fig1, echo=FALSE, fig.cap="\\label{fig:HD_game}Hawk-dove game. Dashed line is the  game theoretical prediction for frequency of hawks.", out.width = '80%'}
knitr::include_graphics("Simulations/mutType_/basicHawkDove.png")
```

@@

To complete the receiver side of the signalling system, we introduce a new 
strategy in the game that we call _learner_. _Learners_ use a mixed strategy in 
the hawk-dove game. Which means, they play hawk (dove) with a certain probability.
Furthermore, the probability with which the play each strategy changes during their lifetime according to the payoff they recieve in previous interactions. In other words, _learners_ learn to behave in the game. To implement the learning
process, we use the reinforcement learning formalism. Specifically, we use the actor-critic algorithm [@sutton_reinforcement_2018]. Individuals ( _learners_ ) perceive the _badge_ of their counterpart as the environmental state in which they find themselves. They estimate the value of that state and a preference for playing hawk (dove) in that state. They update their estimation and preference after each interaction they have. Given that the badge size is a real number between 0 and 1, there are infinitely many environmental states. Thus, for their value estimation and preference computation, individual must use a form of generalization. We let them
use as generalization a function approximation method based on radial basis functions. Each individual uses five _centers_ spread uniformly throughout the range of badge size variation (from 0 to 1). Each time an individual interacts with a conspecific the size of the badge of the interacting partner triggers are response from each one of the _centers_. This response decreases in strength as the badge size is further from the _center_. The decrease follows a radial basis function with gaussian shape. The sum of the responses given by all the centers provide the value estimation and the preference for one of the two pure strategies (hawk or dove). During the learning process individuals update the _weight_ of the response that each one of these centers triggers. In the case of the preference for the strategies the numerical value resulting from the sum of the responses is translated to a probability of one of the two pure strategies through a logistic function. Figure \ref{fig:learning_cartoonRBF} shows a made up example of how the function approximation works for the actor (top panel - probability of choosing dove) and the critic (bottom panel - expected payoff from facing an individual with a given badge size).

@@

In figure \ref{fig:learning_var0.1} we show results of phenotypic and interaction types frequencies from simulations in which individuals vary in their quality. 
The value of quality is drawn from a truncated normal distribution with mean 0.5 
and standard deviation 0.1. Compared to the expectation from evolutionary dynamics of the standard game, the learning scenario with qualiy variation shows lower levels of hawks and therefore lower levels of escalated conflict. At this point individuals do not respond adaptively to the quality of their partner, instead their preference for a given action is dependent on their own quality.

@@

In figure \ref{fig:learning_all_var} we show how the effect of increasing 
variance in quality further deacrases the amount of individuals in a population 
expressing the hawk phenotype.

@@

## How does learning look like in the Hawk-Dove game?

Figures \ref{fig:learning_var0.1} and \ref{fig:learning_all_var} only show how learners behave at a population level, but they do not show how inter-invidivual variation behind those frequencies. In figure \ref{fig:AC_HD} we show the result of the function aproximation mechanism arising at the end of a generation for a sample of the individuals of the populations. These results are show under three different levels of variation in quality. All three panes show that through their learning processes individuals split up in two groups, one which mostly behaves like a hawk and a second that mostly behaves like a dove. The group of "hawks" is mostly, but no exclusively, composed of individuals whose quality is on the upper half of the quality distribution. Concordantly, the group of "doves" is mostly composed of individuals whose qualiy is on the lower half of the quality distribution. The relation between quality and the prefered behaviour of individuals obviously only holds in scenarios with variance in quality. In the scenario with quality invariance, the split of individuals in two groups with a tenedency for each behaviour nevertheless remains.

@@

In figure \ref{fig:learn_dyn} we show the learning trajectory, for the actor and the critic, of a sample of individuals. These dynamics show how individuals split in the two groups according to their experiences. Despite the robustness of the bimodal distribution along the learning dynamics, it is not uncommon that individual go from one group to the other.

@@

Finally, in a new set of simulations we explore the question of how competitive learners are in competition with the pure strategies. Figure \ref{fig:types_comp}
show the results of competition among the three strategies. The population starts with a 1:1 ratio of pure hawks anf doves, and learners are intruced by mutation. Learners manage to increase in frequency. They, however, do not manage to outcompete the pure streategies. In the end, there is a polymorphic equilibrium with the three strategies coexisting, with learners being the most frequen of the threee.

@@

# Cited literature

### 2019-07-01 QD_report.tex
Changes in the text of the Rmd, and adding the bibtex file with references.

\usepackage{float}

\floatplacement{figure}{H}

@@

\hypertarget{the-hawk-dove-game}{%
\section{The Hawk-Dove game}\label{the-hawk-dove-game}}

@@

\begin{figure}
\includegraphics[width=0.8\linewidth]{Simulations//mutType_//basicHawkDove} \caption{\label{fig:HD_game}Hawk-dove game. Dashed line is the  game theoretical prediction for frequency of hawks.}\label{fig:fig1}
\end{figure}

\hypertarget{the-effect-of-learning}{%
\section{The effect of learning}\label{the-effect-of-learning}}

We let individuals have a \emph{badge of status} that potentially
signals their quality. To allow for this signalling, we let the size of
the \emph{badge of status} of an individual be dependent on her quality
through a reaction norm. we assume the size of the \emph{badge} is a
value from 0 to 1. This value is given by \begin{equation}
s_i = \frac{1}{1+e^{-(\alpha_i+\gamma_iQ_i)}}
\label{eq:react_norm}
\end{equation} The parameters \(\alpha\) and \(\gamma\) in equation
\ref{eq:react_norm} determine the shape of the reaction norm, and hence,
how honest the \emph{badge} is in signalling the quality of the
individual.Eventually, the aim is to allow the signalling system to
evolve, by letting the parameters of the reaction norm be inherited from
parent to offspring and be change by a mutational process.

\hypertarget{how-does-learning-work}{%
\subsection{How does learning work?}\label{how-does-learning-work}}

To complete the receiver side of the signalling system, we introduce a
new strategy in the game that we call \emph{learner}. \emph{Learners}
use a mixed strategy in the hawk-dove game. Which means, they play hawk
(dove) with a certain probability. Furthermore, the probability with
which the play each strategy changes during their lifetime according to
the payoff they recieve in previous interactions. In other words,
\emph{learners} learn to behave in the game. To implement the learning
process, we use the reinforcement learning formalism. Specifically, we
use the actor-critic algorithm. Individuals ( \emph{learners} ) perceive
the \emph{badge} of their counterpart as the environmental state in
which they find themselves. They estimate the value of that state and a
preference for playing hawk (dove) in that state. They update their
estimation and preference after each interaction they have. Given that
the badge size is a real number between 0 and 1, there are infinitely
many environmental states. Thus, for their value estimation and
preference computation, individual must use a form of generalization. We
let them use as generalization a function approximation method based on
radial basis functions. Each individual uses five \emph{centers} spread
uniformly throughout the range of badge size variation (from 0 to 1).
Each time an individual interacts with a conspecific the size of the
badge of the interacting partner triggers are response from each one of
the \emph{centers}. This response decreases in strength as the badge
size is further from the \emph{center}. The decrease follows a radial
basis function with gaussian shape. The sum of the responses given by
all the centers provide the value estimation and the preference for one
of the two pure strategies (hawk or dove). During the learning process
individuals update the response that each one of these centers triggers.
In the case of the preference for the strategies the numerical value
resulting from the sum of the responses is translated to a probability
of one of the two pure strategies through a logistic function. Figure
\ref{fig:learning_cartoonRBF} shows a made up example of how the
function approximation works for the actor (top panel - probability of
choosing dove) and the critic (bottom panel - expected payoff from
facing an individual with a given badge size).

@@

\hypertarget{results-when-individuals-do-not-vary-in-their-quality}{%
\subsection{Results when individuals do NOT vary in their
quality}\label{results-when-individuals-do-not-vary-in-their-quality}}

In Figure \ref{fig:learning_invar} we show the phenotypic frequencies
and interaction-type frequencies in 4 generations of a population
composed of only \emph{learners} that do not vary in their quality. The
population is also monomorphic for the reaction norm parameters
(\(\alpha=0\) and \(\beta=0\)). With the values chosen the badge size
distribution is the same as the quality distribution. Thus, the four
generations respresent just replicates of the learning process. The
results show that under quality invariance, the learning algorithm
develops similar phenotypic frequencies as the evolutionary dynamics of
the pure strategies.

@@

\hypertarget{results-when-individuals-vary-in-their-quality}{%
\subsection{Results when individuals vary in their
quality}\label{results-when-individuals-vary-in-their-quality}}

@@

\hypertarget{overal-effect-of-variance}{%
\subsection{Overal effect of variance}\label{overal-effect-of-variance}}

@@

\hypertarget{how-does-learning-look-like-in-the-hawk-dove-game}{%
\subsection{How does learning look like in the Hawk-Dove
game}\label{how-does-learning-look-like-in-the-hawk-dove-game}}

@@

\hypertarget{how-do-the-learning-dynamics-look-like}{%
\subsection{How do the learning dynamics look
like}\label{how-do-the-learning-dynamics-look-like}}

@@

\hypertarget{how-do-learners-fare-against-the-pure-strategies}{%
\section{How do learners fare against the ``pure''
strategies}\label{how-do-learners-fare-against-the-pure-strategies}}

@@

\hypertarget{whats-next}{%
\section{What's next?}\label{whats-next}}

@@

>>>>

\usepackage{float} \floatplacement{figure}{H}

@@

\section{The Hawk-Dove game}\label{the-hawk-dove-game}

@@

\begin{figure}
\includegraphics[width=0.8\linewidth]{Simulations/mutType_/basicHawkDove} \caption{\label{fig:HD_game}Hawk-dove game. Dashed line is the  game theoretical prediction for frequency of hawks.}\label{fig:fig1}
\end{figure}

\section{The effect of learning}\label{the-effect-of-learning}

We let individuals have a \emph{badge of status} that potentially
signals their quality. To allow for this signalling, we let the size of
the \emph{badge of status} of an individual be dependent on her quality
through a reaction norm. we assume the size of the \emph{badge} is a
value from 0 to 1. This value is given by

\begin{equation}
s_i = \frac{1}{1+e^{-(\alpha_i+\gamma_iQ_i)}}
\label{eq:react_norm}
\end{equation}

The parameters \(\alpha\) and \(\gamma\) in equation \ref{eq:react_norm}
determine the shape of the reaction norm, and hence, how honest the
\emph{badge} is in signalling the quality of the individual.Eventually,
the aim is to allow the signalling system to evolve, by letting the
parameters of the reaction norm be inherited from parent to offspring
and be change by a mutational process.

\subsection{How does learning work?}\label{how-does-learning-work}

To complete the receiver side of the signalling system, we introduce a
new strategy in the game that we call \emph{learner}. \emph{Learners}
use a mixed strategy in the hawk-dove game. Which means, they play hawk
(dove) with a certain probability. Furthermore, the probability with
which the play each strategy changes during their lifetime according to
the payoff they recieve in previous interactions. In other words,
\emph{learners} learn to behave in the game. To implement the learning
process, we use the reinforcement learning formalism. Specifically, we
use the actor-critic algorithm Sutton and Barto (2018). Individuals (
\emph{learners} ) perceive the \emph{badge} of their counterpart as the
environmental state in which they find themselves. They estimate the
value of that state and a preference for playing hawk (dove) in that
state. They update their estimation and preference after each
interaction they have. Given that the badge size is a real number
between 0 and 1, there are infinitely many environmental states. Thus,
for their value estimation and preference computation, individual must
use a form of generalization. We let them use as generalization a
function approximation method based on radial basis functions. Each
individual uses five \emph{centers} spread uniformly throughout the
range of badge size variation (from 0 to 1). Each time an individual
interacts with a conspecific the size of the badge of the interacting
partner triggers are response from each one of the \emph{centers}. This
response decreases in strength as the badge size is further from the
\emph{center}. The decrease follows a radial basis function with
gaussian shape. The sum of the responses given by all the centers
provide the value estimation and the preference for one of the two pure
strategies (hawk or dove). During the learning process individuals
update the response that each one of these centers triggers. In the case
of the preference for the strategies the numerical value resulting from
the sum of the responses is translated to a probability of one of the
two pure strategies through a logistic function. Figure
\ref{fig:learning_cartoonRBF} shows a made up example of how the
function approximation works for the actor (top panel - probability of
choosing dove) and the critic (bottom panel - expected payoff from
facing an individual with a given badge size).

@@

\subsection{Results when individuals do NOT vary in their
quality}\label{results-when-individuals-do-not-vary-in-their-quality}

In Figure \ref{fig:learning_invar} we show the phenotypic frequencies
and interaction-type frequencies in 4 generations of a population
composed of only \emph{learners} that do not vary in their quality. The
population is also monomorphic for the reaction norm parameters
(\(\alpha=0\) and \(\gamma=0\)). With the values chosen the badge size
distribution is the same as the quality distribution. Thus, the four
generations respresent just replicates of the learning process. The
results show that under quality invariance, the learning algorithm
develops similar phenotypic frequencies as the evolutionary dynamics of
the pure strategies.

@@

\subsection{Results when individuals vary in their
quality}\label{results-when-individuals-vary-in-their-quality}

@@

\subsection{Overal effect of variance}\label{overal-effect-of-variance}

@@

\subsection{How does learning look like in the Hawk-Dove
game}\label{how-does-learning-look-like-in-the-hawk-dove-game}

@@

\subsection{How do the learning dynamics look
like}\label{how-do-the-learning-dynamics-look-like}

@@

\section{\texorpdfstring{How do learners fare against the ``pure''
strategies}{How do learners fare against the pure strategies}}\label{how-do-learners-fare-against-the-pure-strategies}

@@

\section{What's next?}\label{whats-next}

@@

\section*{Cited literature}\label{cited-literature}
\addcontentsline{toc}{section}{Cited literature}

\hypertarget{refs}{}
\hypertarget{ref-sutton_reinforcement_2018}{}
Sutton, Richard S., and Andrew G. Barto. 2018. \emph{Reinforcement
Learning: An Introduction}. Edited by Francis Bach. Second edition
edition. Cambridge, MA: A Bradford Book.
